\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-25 paper.qmd
Updated projections with tSNE

<!-- UMAP layout with author's suggested parameter choice-->
```{r}
#| label: published-pbmc
## Import data
training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50.rds")
training_data_pbmc <- training_data_pbmc[, 1:9] |>
  mutate(ID = 1:NROW(training_data_pbmc))

@@

<!--Fit the best model for author suggestion and compute error-->
```{r}
#| label: hexbin-pbmc
## Compute hexbin parameters
num_bins_x_pbmc <- 30
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = umap_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "PC_"
)

@@

trimesh_removed_pbmc <- ggplot() + 
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#33a02c") +
  geom_point(
    data = umap_pbmc_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.3,
    size = 0.1,
    color = "#000000"
  )  +
  interior_annotation("a")

@@

```{r}
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
trimesh_removed_pbmc 
```

::: {#fig-pbmc1-sc layout-ncol="4" fig-pos="H"}
![](figures/pbmc3k/umap_trimesh_plot.png)

![](figures/pbmc3k/sc_1.png)

![](figures/pbmc3k/sc_2.png)

![](figures/pbmc3k/sc_3.png)

Model in \gD{}, on the UMAP layout, and three views of the fit in projections from $9\text{-}D$, for the PBMC3k data ($(s_1, \ s_2) = (-0.050, \ -0.041)$, $b = 870 \  (30, \ 29)$, $m = 135$, benchmark value to remove large edges is $0.099$) (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/VqqWuE0Jj6A>).
:::

@@

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = tsne_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.05,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "PC_"
)

@@

tr_from_to_df_pbmc <- tr_from_to_df_pbmc |>
  filter(row_number() != 360)

@@

## Compute error
error_df <- augment(
  df_bin_centroids = df_bin_centroids_pbmc,
  df_bin = df_bin_pbmc,
  training_data = training_data_pbmc,
  newdata = NULL,
  type_NLDR = "tSNE",
  col_start = "PC_")

@@

## Hexagonal binning to have regular hexagons
hb_obj_pbmc <- hex_binning(
  data = tsne_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.05)

@@

trimesh_removed_mnist <- ggplot() + 
  geom_segment(data = tr_from_to_df_mnist, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#33a02c") +
  geom_point(data = pacmap_minst_scaled,
             aes(
               x = PaCMAP1,
               y = PaCMAP2
             ),
             alpha=0.5, 
             size = 0.1) +
  interior_annotation("a", sc_ltr_pos_mnist)

>>>>

<!-- UMAP layout with author's suggested parameter choice-->
```{r}
#| label: published-pbmc
## Import data
training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50.rds")
names(training_data_pbmc) <- paste0("PC", 1:50)

training_data_pbmc <- training_data_pbmc[, 1:9] |>
  mutate(ID = 1:NROW(training_data_pbmc))

@@

<!--Fit the best model for author suggestion and compute error-->
```{r}
#| label: hexbin-pbmc
## Compute hexbin parameters
num_bins_x_pbmc <- 26
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = umap_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "PC"
)

@@

trimesh_removed_pbmc <- ggplot() + 
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#33a02c") +
  geom_point(
    data = umap_pbmc_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    alpha = 0.3,
    size = 0.1,
    color = "#000000"
  )  +
  coord_fixed() +
  interior_annotation("a")

@@

```{r}
#| label: pbmc-umap-model-proj

# Apply the scaling
scaled_pbmc_data <- scale_data_manual(training_data_pbmc |> select(-ID)) |>
  as_tibble()

df_b_pbmc <- df_bin_pbmc |>
  dplyr::filter(hb_id %in% df_bin_centroids_pbmc$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_pbmc <- df_b_pbmc[match(df_bin_centroids_pbmc$hexID, df_b_pbmc$hb_id),] |>
  dplyr::select(-hb_id) |>
  select(-type)

# Apply the scaling
scaled_pbmc_data_model <- scale_data_manual(df_b_pbmc) |>
  as_tibble()

# Combine with the true model for visualization
df <- dplyr::bind_rows(scaled_pbmc_data_model |> mutate(type = "model"),
                       scaled_pbmc_data |> mutate(type = "data"))

## First projection
projection <- cbind(
    c(0.24156,0.85058,0.11564,-0.03255,-0.28106,0.27463,-0.10539,0.08429,0.13107),
    c(-0.67212,0.02745,-0.26718,-0.33126,-0.34919,0.20839,-0.14498,-0.37708,0.15455))

projected <- as.matrix(scaled_pbmc_data) %*% projection

projected_model <- as.matrix(scaled_pbmc_data_model) %*% projection

model_df <- dplyr::left_join(
  distance_df_small_edges_pbmc |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

axes_obj <- gen_axes(
  data = scaled_pbmc_data, 
  proj_data = projected_df, 
  limits = 1)

pbmc_proj_tsne_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  interior_annotation("b")

## Second projection
projection <- cbind(
    c(0.78689,-0.14774,0.36527,0.09617,0.12097,-0.00033,0.36973,-0.20959,-0.08438),
    c(0.18390,-0.43121,-0.30398,0.05861,0.55282,-0.14058,-0.27048,0.44386,-0.27372))

projected <- as.matrix(scaled_pbmc_data) %*% projection

projected_model <- as.matrix(scaled_pbmc_data_model) %*% projection

model_df <- dplyr::left_join(
  distance_df_small_edges_pbmc |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

axes_obj <- gen_axes(
  data = scaled_pbmc_data, 
  proj_data = projected_df, 
  limits = 1)

pbmc_proj_tsne_model2 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  interior_annotation("c")

## Third projection
projection <- cbind(
    c(0.00067,0.53963,-0.16241,-0.09777,0.32766,0.44543,-0.51359,-0.29884,0.00951),
    c(0.62399,0.26016,0.06441,-0.34686,-0.40466,0.24643,0.23537,0.07972,0.34402))

projected <- as.matrix(scaled_pbmc_data) %*% projection

projected_model <- as.matrix(scaled_pbmc_data_model) %*% projection

model_df <- dplyr::left_join(
  distance_df_small_edges_pbmc |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

axes_obj <- gen_axes(
  data = scaled_pbmc_data, 
  proj_data = projected_df, 
  limits = 1)

pbmc_proj_tsne_model3 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  interior_annotation("d")

```{r}
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-height: 10
#| fig-width: 10
#| fig-cap: "Model in \\gD{}, on the UMAP layout, and three views of the fit in projections from $9\\text{-}D$, for the PBMC3k data ($(s_1, \ s_2) = (-0.050, \ -0.041)$, $b = 870 \  (30, \ 29)$, $m = 135$, benchmark value to remove large edges is $0.099$) (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/VqqWuE0Jj6A>)."

trimesh_removed_pbmc + pbmc_proj_tsne_model1 +
  pbmc_proj_tsne_model2 + pbmc_proj_tsne_model3 +
  plot_layout(ncol = 2)
```

<!-- ::: {#fig-pbmc1-sc layout-ncol="4" fig-pos="H"} -->
<!-- ![](figures/pbmc3k/umap_trimesh_plot.png) -->

<!-- ![](figures/pbmc3k/sc_1.png) -->

<!-- ![](figures/pbmc3k/sc_2.png) -->

<!-- ![](figures/pbmc3k/sc_3.png) -->

<!-- Model in \gD{}, on the UMAP layout, and three views of the fit in projections from $9\text{-}D$, for the PBMC3k data ($(s_1, \ s_2) = (-0.050, \ -0.041)$, $b = 870 \  (30, \ 29)$, $m = 135$, benchmark value to remove large edges is $0.099$) (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/VqqWuE0Jj6A>). -->
<!-- ::: -->

@@

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = tsne_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "PC"
)

@@

# tr_from_to_df_pbmc <- tr_from_to_df_pbmc |>
#   filter(row_number() != 360)

@@

## Compute error
error_df <- augment(
  df_bin_centroids = df_bin_centroids_pbmc,
  df_bin = df_bin_pbmc,
  training_data = training_data_pbmc,
  newdata = NULL,
  type_NLDR = "tSNE",
  col_start = "PC")

@@

## Hexagonal binning to have regular hexagons
hb_obj_pbmc <- hex_binning(
  data = tsne_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1)

@@

trimesh_removed_mnist <- ggplot() + 
  geom_segment(data = tr_from_to_df_mnist, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#33a02c") +
  geom_point(data = pacmap_minst_scaled,
             aes(
               x = PaCMAP1,
               y = PaCMAP2
             ),
             alpha=0.5, 
             size = 0.1) +
  coord_fixed() +
  interior_annotation("a", sc_ltr_pos_mnist)

### 2024-08-25 paper.tex
Updated trimesh for pbmc tSNE model
Updated projections with tSNE

To determine whether the UMAP representation with the (hyper-)parameter
choice suggested by \citet{chen2023} preserves the original data
structure, we visualize the model constructed with UMAP overlaid on the
\(9\text{-}D\) data. The \textbf{?@fig-umap-author} shows three
well-separated clusters. However, as shown in Figure~\ref{fig-pbmc1-sc},
there is no big separation between three clusters in \(9\text{-}D\).
Therefore, the suggested UMAP representation
(\textbf{?@fig-umap-author}) does not accurately represent the
structure(s) present in PBMC3k dataset. But, when visualizing the
model-in-th-data-space some unobserved structures can be seen. Some
clusters have non-linear continuity patterns and high-density patches
(Figure~\ref{fig-pbmc1-sc}).

@@

\begin{minipage}{0.25\linewidth}
\includegraphics{figures/pbmc3k/umap_trimesh_plot.png}\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
\includegraphics{figures/pbmc3k/sc_1.png}\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
\includegraphics{figures/pbmc3k/sc_2.png}\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
\includegraphics{figures/pbmc3k/sc_3.png}\end{minipage}%

\caption{\label{fig-pbmc1-sc}Model in \gD{}, on the UMAP layout, and
three views of the fit in projections from \(9\text{-}D\), for the
PBMC3k data (\((s_1, \ s_2) = (-0.050, \ -0.041)\),
\(b = 870 \  (30, \ 29)\), \(m = 135\), benchmark value to remove large
edges is \(0.099\)) (The \textbf{langevitour} software is used to view
the data with a tour, and the full video is available at
\url{https://youtu.be/VqqWuE0Jj6A}).}

>>>>

To determine whether the UMAP representation with the (hyper-)parameter
choice suggested by \citet{chen2023} preserves the original data
structure, we visualize the model constructed with UMAP overlaid on the
\(9\text{-}D\) data. The \textbf{?@fig-umap-author} shows three
well-separated clusters. However, as shown in \textbf{?@fig-pbmc1-sc},
there is no big separation between three clusters in \(9\text{-}D\).
Therefore, the suggested UMAP representation
(\textbf{?@fig-umap-author}) does not accurately represent the
structure(s) present in PBMC3k dataset. But, when visualizing the
model-in-th-data-space some unobserved structures can be seen. Some
clusters have non-linear continuity patterns and high-density patches
(\textbf{?@fig-pbmc1-sc}).

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-model-pbmc-author-proj-1.pdf}

\caption{\label{fig-model-pbmc-author-proj}Model in \gD{}, on the UMAP
layout, and three views of the fit in projections from \(9\text{-}D\),
for the PBMC3k data (\((s_1,  s_2) = (-0.050,  -0.041)\),
\(b = 870   (30,  29)\), \(m = 135\), benchmark value to remove large
edges is \(0.099\)) (The \textbf{langevitour} software is used to view
the data with a tour, and the full video is available at
\url{https://youtu.be/VqqWuE0Jj6A}).}
